Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-30 doc/ApTranscriptome_manuscript.Rmd
Incorporated comments from AE/RD/SCH/AN to abstract

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species, their abundances, and underlies their future responses to climate change. However, the genetic basis of thermal tolerance is poorly understood for most species. In this study, we present a novel regression approach to characterize the thermal reaction norm for all genes in an organism's transcriptome. We apply this approach to parapatric ant species in the eastern United States; *A. carolinensis* and *A. picea*, which have warmer and colder distributions, respectively. Using this approach, we find that at least 10% of genes are thermally responsive. In both species, more genes are up-regulated in response to cold than warm temperatures, suggesting that there is less adaptive potential to heat than cold stress. Between species, *A. carolinensis* shuts down more genes in response to thermal extremes, whereas *A. picea* activates more genes in response to cold and hot temperatures. The genes that are shut down in *A. carolinensis* but up-regulated in *A. picea* included many metabolism, cellular organization and repair gene ontology terms. We conclude that these species have different mechanisms for responding to thermal stress. *A. picea* resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. In summary, this study demonstrates how we can gain insight on the genetic basis of local adaption to climate using a reaction norm approach with transcriptomic data.

>>>>

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew D. Nguyen, Mahesh Vangala, Jim Vincent, Robert R. Dunn,
  Aaron Ellison, Nate Sanders, Nicholas J. Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. However, the full set of genes and pathways underlying thermal tolerance are poorly understood for most species. Here, we demonstrate how insights about the genetic basis of local adaption to climate can be gained by applying a reaction norm approach to transcriptomic data. We introduce the thermal *reactionome*, the characterization of the thermal reaction norm of all genes in an organism's transcriptome. We characterized transcriptome-wide gene expression over a 38-degree thermal range for two closely-related ant species, *A. carolinensis* and *A. picea*, which are parapatrically distributed along a temperature gradient in eastern deciduous forests of North America. At least 10% of genes are thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, \todo{a lower magnitude of upregulation at thermal extremes}, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in \todo{metabolism, cellular organization and repair}. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.
